Fractional Types

### 2013-02-16 frac.tex
subcoded going crazy
another pass at intro with citations etc

\begin{document}
\title{Fractional Types}
\author{Roshan P. James \and Zachary Sparks \and Jacques Carette \and Amr Sabry}
\institute{}
\maketitle

@@

In the physics community, Landauer~\cite{Landauer:1961}, Feynman, and others
have interpreted the laws of physics as fundamentally related to
computation. In more detail, the great majority of the laws of physics are
formulated as equalities between different physical observables. For
instance, Newton's second law of classical mechanics equates the force acting
on a system to its rate of change of momentum. The insight of Landauer,
Feynman, etc. is that \emph{different} physical observables should not be
related by an \emph{equality} but more appropriately by an \emph{isomorphism}
that witnesses, explains, and models the process of transforming one
observable to the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive mathematics. This has been further
extended (first by Streicher and Awodey, and subsequently by many others)
with a new \emph{identity type} representing the proposition
that two terms are ``equal.'' Specifically, given two terms $a$ and $b$ of
the same type $A$, one forms the type $\texttt{Id}_A(a,b)$ representing the
proposition that $a$ and $b$ are equal: in other words, a term of type
$\texttt{Id}_A(a,b)$ witnesses, explains, and models the process of
transforming $a$ to $b$ and vice-versa.

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function. More recently, type isomorphisms have taken a more central role as
\emph{the} fundamental computational mechanism from which more conventional,
i.e., irreversible computation, is derived~\cite{infeffects}. In more detail,
in our own previous work, we started with the notion of type isomorphism as
the ``right'' starting point for a theory of computation. The central
technical contribution was the development of a family of programming
languages, {{Pi}} with various superscripts, in which every computation
is an isomorphism. Other major results are that all such computations
preserve the information-theoretic entropy, that the language is logically
reversible, and that it has a model in symmetric monoidal categories (which
are the backbone of models for linear logic and quantum computation). We have
also invested a significant amount of time in using {{Pi}} as an actual
programming language and established several idioms for programming large
classes of interesting (recursive) programs in {{Pi}}.

@@

\roshan{It is unclear what the Id types refer to. We should provide
  some more context here}

This paper addresses this higher-order extension.  It does this via a novel
type construction, which we dub a \emph{fractional type} because of its
interaction with product types.  By adding a formal (linear) dual to
(type theoretic) products, we can internalize the notion that ``taking an
input'' corresponds to negative information, which turns out to be the 
basis of first-class functions.

The plan of the paper is therefore
as follows: \todo{roadmap}

@@

In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain
the simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
richer languages, e.g., in a Turing complete language that admits
non-termination, isomorphisms necessarily denote \emph{partial
  bijections}~\cite{rc2011,infeffects}. As we demonstrate in this section,
once we make the extension to higher-order functions, we must interpret the
syntactic isomorphisms as \emph{relations}. The intuitive idea, is that the
fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be able to map the
element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.

>>>>

\begin{document}
\title{Fractional Types}
\author{Roshan P. James \and Zachary Sparks \and Jacques Carette \and Amr Sabry}
\institute{Indiana University and McMaster University}
\maketitle

@@

In the physics community, Landauer~\cite{Landauer,Landauer:1961},
Feynman~\cite{springerlink:10.1007/BF02650179}, and others have interpreted
the laws of physics as fundamentally related to computation. In more detail,
the great majority of the laws of physics are formulated as equalities
between different physical observables. For instance, Newton's second law of
classical mechanics equates the force acting on a system to its rate of
change of momentum. The insight of Landauer, Feynman, etc. is that
\emph{different} physical observables should not be related by an
\emph{equality} but more appropriately by an \emph{isomorphism} that
witnesses, explains, and models the process of transforming one observable to
the other.

In the mathematics and logic community, Martin-L\"of developed an extension
of the simply typed $\lambda$-calculus originally intended to provide a
rigorous framework for constructive
mathematics~\cite{citeulike:7374951}. This theory has been further extended
with \emph{identity types} representing the proposition that two terms are
``equal.'' (See~\cite{warren,streicher} for a survey.) Briefly speaking,
given two terms $a$ and $b$ of the same type $A$, one forms the type
$\texttt{Id}_A(a,b)$ representing the proposition that $a$ and $b$ are equal:
in other words, a term of type $\texttt{Id}_A(a,b)$ witnesses, explains, and
models the process of transforming $a$ to $b$ and vice-versa.

In the computer science community, the theory and practice of type
isomorphisms is well-established. Originally, such type isomorphisms were
motivated by the pragmatic concern of searching large libraries of functions
by providing one of the many possible isomorphic types for the desired
function~\cite{Rittri:1989:UTS:99370.99384}. More recently, type isomorphisms
have taken a more central role as \emph{the} fundamental computational
mechanism from which more conventional, i.e., irreversible computation, is
derived. In more detail, in our own previous
work~\cite{James:2012:IE:2103656.2103667,rc2011,rc2012} we started with the
notion of type isomorphism as the ``right'' starting point for a theory of
computation. The central technical contribution was the development of a
family of programming languages, {{Pi}} with various superscripts, in which
every computation is an isomorphism. Other major results are that all such
computations preserve the information-theoretic entropy, that the languages
are logically reversible, and that they have models in various flavors of
symmetric monoidal categories (which are the backbone of models for linear
logic and quantum computation). We have also invested a significant amount of
time in using {{Pi}} as an actual programming language and established
several idioms for programming large classes of interesting (recursive)
programs in {{Pi}}.

@@

Technically speaking, obtaining a higher-order extension requires the
construction of a \emph{compact closed category} from the underlying monoidal
category for {{Pi}}. Although several such constructions that work in special
cases are known in the literature (surveyed below), there are many subtleties
in adapting these ideas to {{Pi}}. Our main novel technical device to
achieving the higher-order extension is a \emph{fractional type}, which we so
named because of its duality with conventional product types.  By adding a
formal (linear) dual to (type theoretic) products, we can internalize the
notion that ``taking an input'' corresponds to negative information, which
turns out to be the basis of first-class functions. The remainder of the
paper reviews {{Pi}} and then introduces the syntax and semantics of the
extension with fractional types. We then study the properties of the extended
language and establish its expressiveness via several constructions and
examples that exploit its ability to model higher-order computations.

%% \roshan{It is unclear what the Id types refer to. We should provide
%%   some more context here}
%% I think the previous paragraph in the intro that talks about Martin-Lof type 
%% theory provides the context. Anything else??

@@

In {{Pi}} presented in Sec.~\ref{sec:pi}, we have been able to maintain the
simple idea that a syntactic isomorphism denotes a \emph{bijection}. In
richer languages, e.g., in a Turing complete language that admits
non-termination, isomorphisms necessarily denote \emph{partial
  bijections}~\cite{rc2011,James:2012:IE:2103656.2103667}. As we demonstrate
in this section, once we make the extension to higher-order functions, we
must interpret the syntactic isomorphisms as \emph{relations}. The intuitive
idea, is that the fission point combinator {{eta*_b : 1 <-> 1/b * b}} must be
able to map the element {{() : 1}} to any pair of the form {{(1/v,v)}} with
{{v:b}}. Similarly, the combinator {{eps*_b : 1/b * b <-> 1}} must be able to
accept pairs {{(1/v1,v2)}} in which the negative information {{1/v1}} does
not match the positive information {{v2}}. On such pairs, {{eps*_b}} is
undefined, i.e., it does not produce {{():1}}. In other words, some of the
combinators are not partial and some produce more than one result for a
particular input, and hence we move to the setting of general binary
relations to interpret our combinators.
